---
title: "Atlas Variables Calculations"
author: "Friederike Wölke"
date: "2024-04-20"
output:
  html_document:
    code_folding: hide
format:
  html:
    toc: true
    number-sections: true
    colorlinks: true
---

# Clean Environment
```{r, message = F}
rm(list=ls())
gc()
```

# Libraries
```{r, mesaage = F}
# Spatial:
library(sf) 
sf_use_s2(FALSE) # switch spherical geometry off

# Data handling:
library(rstatix)
library(dplyr) 
library(tidyr) 
# library(plyr) 
## plyr is required but produces some issues with dplyr. Thus it will be called in each function individually and is not required to load here.

# Telfer calculation:
## sparta is not on CRAN and cannot be updated as usual because there is another package called sparta from CRAN but we don't want this.
# library(devtools)
# install_github('biologicalrecordscentre/sparta')
library(sparta) 

```

# Load Environment
```{r}
load("../../out/RData/1_Atlas_Data_clean.RData")
ls()
```


# Calculations
## 1. Occupancy
### 1.1 Calculate Areas from Atlases
```{r,mesaage = F, warning = F}
pres_dat_full_grain <- list()
pres_dat_full_atlas <- list()

for (n_atlas in seq_along(atlas_names)){
  atlas_name <- atlas_names[n_atlas] # set variable for current run of the loop
  atlas <- grid_list[[n_atlas]] # subset list of atlases to one
  
  for (grain in seq_along(atlas)){
    # set some variables for current run of the loop
    grain_a <- names(atlas)[grain]
    grain_d <- as.numeric(gsub("\\D", "", grain_a)) # "\\D" = remove all non-numbers
    
    # subset atlas to one grain
    atlas_1scale <- atlas[[grain]] %>% 
      select(cell_grouping, cell_label, area, cell_long, cell_lat) %>%
      st_drop_geometry()


    ## =============================================================== ###
    ## ================= Calculate Areas ============================= ###
    ## =============================================================== ###


    # Total Area and Total number of cells of the Atlas (as measured from grid)
    atlas_1scale <- atlas_1scale %>% 
      group_by(cell_grouping) %>%
      mutate(
        Total_area = sum(atlas_1scale$area),
        Total_Ncells = n_distinct(cell_label))
    
    # Total number of sampled cells and total Area sampled per sampling period
    Total_Ncells_samp <- presence_data_filt %>% 
      filter(dataset == atlas_name & cell_grouping == grain_d) %>%
      select(dataset, cell_grouping, cell_label, area) %>% unique() %>% na.omit() %>%
      group_by(dataset, cell_grouping) %>% distinct(cell_label, .keep_all = T) %>%
      summarise(
        n_cells = n_distinct(cell_label),
        area_samp = sum(area)) %>% 
      ungroup() %>%
      distinct(n_cells, area_samp)
    
    # Add columns to atlas
    atlas_1scale$Total_area_samp <- Total_Ncells_samp$area_samp
    atlas_1scale$Total_Ncells_samp <- Total_Ncells_samp$n_cells
    map_atlas_1scale <- atlas_1scale
    
    # subset the presence/absence data to the current spatial grain:
    pres_data_1atlas <- presence_data_filt %>% 
      filter(dataset == atlas_name) %>%
      filter(cell_grouping == grain_d) %>%
      select(dataset, tp, verbatim_name, cell_label, cell_grouping)
  
    # Add Areas to data
    pres_data_full_1atlas <- left_join(map_atlas_1scale, pres_data_1atlas)
    pres_data_full_1atlas %>% filter(!is.na(verbatim_name))
    pres_data_full_1atlas$dataset <- atlas_names[n_atlas]
    pres_dat_full_grain[[grain]] <- pres_data_full_1atlas %>% 
        unique() %>% 
        filter(!is.na(verbatim_name))
    
  }
  
  pres_dat_grain2 <- plyr::rbind.fill(pres_dat_full_grain, fill=T)
  pres_dat_full_atlas[[n_atlas]] <- pres_dat_grain2
  
}

# Bind list to dataframe
pres_dat_full <- plyr::rbind.fill(pres_dat_full_atlas, fill=T)
pres_dat_full[pres_dat_full == "<NA>"] = NA #Fix NA issues in the data (takes a while.)

# Add Column for grain size (in km, i.e., approx. length of the cell in the grid)
pres_dat_final <- pres_dat_full %>%  
  mutate(cell_grouping = factor(cell_grouping,levels = desired_levels),
        grain = case_when(
          dataset == atlas_names[1] ~ "10",
          dataset == atlas_names[2] ~ "5",
          dataset == atlas_names[3] ~ "20", 
          dataset == atlas_names[4] ~ "50")) %>% 
  filter(!is.na(verbatim_name))

### =============================================== ###
rm(pres_dat_full_atlas, pres_dat_full_grain, pres_dat_grain2, pres_dat_full, pres_data_1atlas, pres_data_full_1atlas, 
map_atlas_1scale, atlas_1scale,
Total_Ncells_samp, presence_data_filt, atlas, atlas_name, grain, grain_a, grain_d, n_atlas)
### =============================================== ###

```


### 1.2 Calculate Occupancy 
```{r}
occ_data <- pres_dat_final %>%
  ungroup() %>% 
  distinct(dataset, tp, cell_grouping, verbatim_name, cell_label, .keep_all=T) %>%
  group_by(dataset, tp, cell_grouping, verbatim_name) %>% 

# Calculate AOO:
  mutate(
    mean_area = mean(area),
    AOO = sum(area),
    occ_Ncells = n_distinct(cell_label)) %>%

# Calculate relative Occupancy:
  mutate(
    rel_AOO = AOO/Total_area_samp,
    rel_occ_Ncells = occ_Ncells/Total_Ncells_samp) %>% 

  ungroup() %>% group_by(dataset, tp, cell_grouping) %>%

  mutate(
    Gamma = n_distinct(verbatim_name)) %>%
# Remove duplicated rows:
  distinct() 

# create scale column as a fraction of the full country:
occ_data_final <- occ_data %>% 
  ungroup() %>%
  group_by(dataset) %>%
  mutate(
    scale = ifelse(dataset %in% c(atlas_names[1]) & cell_grouping == "1", 1/64, NA),
    scale = ifelse(dataset %in% c(atlas_names[1]) & cell_grouping == "2", 1/32, scale), 
    scale = ifelse(dataset %in% c(atlas_names[1]) & cell_grouping == "4", 1/16, scale), 
    scale = ifelse(dataset %in% c(atlas_names[1]) & cell_grouping == "8", 1/8, scale),
    scale = ifelse(dataset %in% c(atlas_names[1]) & cell_grouping == "16", 1/4, scale),
    scale = ifelse(dataset %in% c(atlas_names[1]) & cell_grouping == "32", 1/2, scale),
    scale = ifelse(dataset %in% c(atlas_names[1]) & cell_grouping == "64", 1, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "1", 1/128, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "2", 1/16, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "4", 1/32, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "8", 1/16, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "16", 1/8, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "32", 1/4, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "64", 1/2, scale),
    scale = ifelse(dataset %in% c(atlas_names[2:4]) & cell_grouping == "128", 1, scale))

colSums(is.na(occ_data_final))


# save reduced version of this to file:

species_data <- occ_data_final %>% 
  select(
  dataset, tp, cell_grouping, verbatim_name, 
  mean_area, Total_area, Total_area_samp, Total_Ncells, Total_Ncells_samp, 
  Gamma,AOO, occ_Ncells, rel_occ_Ncells, rel_AOO) %>%
  distinct(dataset, tp, cell_grouping, verbatim_name, .keep_all = T) 

### =============================================== ###
rm(occ_data, pres_dat_final)
### =============================================== ###

```

## 2. Occupancy-Area-Relationship (OAR)
### 2.1 Mark saturated scales 
- code dummy column "exclude" to exclude species with < 2 scales not saturated (i.e, if less than two scales are available for OAR calculation)
- Saturation: rel_AOO = 1

```{r}
list_sp <- list()
list_tp <- list()
list_a <- list()

for (a in seq_along(atlas_names)){
  temp_df <- species_data %>% filter(dataset == atlas_names[a])

  for (t in seq_along(time_periods)) {
    temp_df_t <- temp_df %>% filter(tp == time_periods[t])
    species_names <- unique(temp_df_t$verbatim_name)

    for (s in seq_along(species_names)){
      temp_df_s <- temp_df_t %>% filter(verbatim_name == species_names[s])
      
      # Exclude saturated scales
      temp_df_red <- temp_df_s %>% filter(rel_AOO < 1)
      
      if (nrow(temp_df_red) < 2 & nrow(temp_df_red) > 0) {
        out_df <- data.frame(verbatim_name = species_names[s],
                             dataset = atlas_names[a],
                             tp = time_periods[t],
                             exclude = 1,
                             available_scales = nrow(temp_df_red),
                             mean_relAOO = mean(temp_df_red$rel_AOO, na.rm=T))
      } else if (nrow(temp_df_red) >= 2) {
        out_df <- data.frame(verbatim_name = species_names[s],
                             dataset = atlas_names[a],
                             tp = time_periods[t],
                             exclude = 0,
                             available_scales = nrow(temp_df_red),
                             mean_relAOO = mean(temp_df_red$rel_AOO, na.rm=T))
      } else {
                out_df <- data.frame(verbatim_name = species_names[s],
                             dataset = atlas_names[a],
                             tp = time_periods[t],
                             exclude = 1,
                             available_scales = nrow(temp_df_red),
                             mean_relAOO = 1)
                }
      list_sp[[s]] <- out_df
    
      }
    sp_df <- plyr::rbind.fill(list_sp)
    list_tp[[t]] <- sp_df
  
    }
  tp_df <- plyr::rbind.fill(list_tp)
  list_a[[a]] <- tp_df

}

atlas_df <- plyr::rbind.fill(list_a)

str(atlas_df)
str(species_data)
sp_data_new <- full_join(species_data, atlas_df) %>% distinct(dataset, tp, cell_grouping, verbatim_name, .keep_all=T)

### =============================================== ###
rm(a,t,s, temp_df, temp_df_red, temp_df_s, temp_df_t, atlas_df, species_data, sp_df, tp_df, list_a, list_sp, list_tp, out_df)
### =============================================== ###

```

### 2.2 Calculate OAR & Fractal Dimension
```{r}
dd <- sp_data_new %>%
  filter(exclude == 0) %>%
  filter(rel_occ_Ncells < 1) %>% # exclude saturated scales
  unique() %>%
  filter_at(
    vars(c(
      cell_grouping, AOO, mean_area)), 
    any_vars(!is.na(.)))

## Variables:
OAR_list_sp <- list()
OAR_list_sp_tp <- list()
OAR_list_sp_tp_atlas <- list()

## Loop:
for (n_atlas in seq_along(atlas_names)){
  atlas_name <- atlas_names[n_atlas]
  atlas <- dd %>% filter(dataset == atlas_name)

  for (time in seq_along(time_periods)){
    period_nr <-  time_periods[time]
    atlas_1time <- atlas %>% filter(tp == period_nr) 
    sp <- unique(atlas_1time$verbatim_name)
    
    for(spec in seq_along(sp)){
      species <- sp[spec]
      model_df <- atlas_1time %>% 
        filter(verbatim_name == species) %>% 
        distinct()

      OAR <- lm(log(AOO) ~ log(mean_area), data = model_df)
      
      OAR_df <- data.frame(
        verbatim_name = species,
        dataset = atlas_name,
        tp = period_nr,
        m_AOO_a = OAR$coefficients[2],
        b_AOO_a = OAR$coefficients[1])
      
      OAR_df$D_AOO_a <- -2*OAR_df$m_AOO_a+2
      OAR_df$dataset <- as.factor(OAR_df$dataset)
      OAR_list_sp[[spec]] <- OAR_df
      
    }
    
    OAR_df_sp <- plyr::rbind.fill(OAR_list_sp, fill = T)
    OAR_list_sp_tp[[time]] <- OAR_df_sp
     
  }
  OAR_df_sp_tp <- plyr::rbind.fill(OAR_list_sp_tp, fill = T)   
  OAR_list_sp_tp_atlas[[n_atlas]] <- OAR_df_sp_tp
}
    
OAR_final <- plyr::rbind.fill(OAR_list_sp_tp_atlas, fill = T) %>% distinct()
colSums(is.na(OAR_final))


species_data_new <- merge(sp_data_new, 
                          OAR_final, 
                          by=c(intersect(names(sp_data_new), names(OAR_final))), all = T) %>% distinct()

### =============================================== ###
rm(dd, OAR_list_sp_tp_atlas, OAR_list_sp_tp, OAR_df_sp_tp, OAR_list_sp, OAR_df_sp, OAR_df, sp, atlas_1time, model_df, atlas_name, OAR_final, n_atlas, OAR, period_nr, spec, species, species_names, time )


colSums(is.na(species_data_new))

species_data_new %>% 
    filter(exclude == 1) %>%    
    group_by(dataset, tp) %>% 
    summarize(n = n_distinct(verbatim_name)) 
### =============================================== ###

```

## 3. Telfer index of relative change

```{r}
# Telfer:
telfer_res <- list()
for (i in seq_along(atlas_names)){
  df <- occ_data_final %>% 
    filter(cell_grouping == 1 & dataset == atlas_names[i]) %>%
    distinct(dataset, tp, verbatim_name, cell_label) 
  telfer <- sparta::telfer(taxa = df$verbatim_name,
                         site = df$cell_label,
                         time_period = df$tp,
                         minSite = 1)
telfer$dataset <- atlas_names[i]
telfer_res[[i]] <- telfer
  
}

telfer_res_df <- plyr::rbind.fill(telfer_res, fill = T) %>% distinct() %>%
  select(taxa, Telfer_1_2, dataset) %>% 
  rename(verbatim_name = taxa)%>%
  reorder_levels(dataset, order = atlas_names)

telfer_res_df$Telfer_1_2 <- round(telfer_res_df$Telfer_1_2, 3)
species_data_new2 <- left_join(species_data_new, telfer_res_df)

### =============================================== ###
colSums(is.na(species_data_new2))
rm(atlas, df, species_data_new, sp_data_new,telfer, telfer_res_df, i, telfer_res )
### =============================================== ###
```

## 4. Log Ratio of Change in Occupancy
### 4.1 Long to wide format

```{r}

wide_dfs <- list()
for (i in seq_along(time_periods)){

    wide_dfs[[i]] <- species_data_new2 %>%
        filter(cell_grouping == 1) %>% 
        distinct(dataset, tp, verbatim_name, AOO) %>%
        group_by(dataset, tp, verbatim_name) %>%
        filter(tp == time_periods[i]) %>%
        setNames(paste0('tp', i, '_', names(.))) %>% 
        ungroup() %>%
        select(-c(paste0("tp", i, "_tp"))) %>%
        rename(
            verbatim_name = paste0("tp", i, "_verbatim_name"),
            dataset = paste0("tp", i, "_dataset"))
       }

sp_dat_wide <- merge(wide_dfs[[1]], wide_dfs[[2]])
colSums(is.na(sp_dat_wide))


### =============================================== ###
rm(wide_dfs, i)
### =============================================== ###
```

## 4.2 Calculate Log Ratio of Occupancy Change


```{r}
# Calculate log-Ratio of AOO (Temporal change)
logRatio <- sp_dat_wide %>% 
  mutate(log_R2_1 = log(tp2_AOO/tp1_AOO)) %>%
  select(-tp1_AOO, -tp2_AOO)

big_table <- full_join(species_data_new2, logRatio) %>% filter(cell_grouping == 1) %>%
    distinct(dataset, tp, cell_grouping, verbatim_name, .keep_all=T) %>%
    mutate_if(is.numeric, round, 3)


big_table %>% group_by(dataset, tp) %>% 
  summarise(across(everything(), ~ sum(is.na(.x))))

### =============================================== ###
rm(logRatio,sp_dat_wide)
### =============================================== ###

```


# Save Output:
```{r}
saveRDS(occ_data_final, "../../out/rds/SpeciesDatPerCell.rds") # data per cell
saveRDS(big_table, "../../out/predictors/BigTable.rds") # data per species
save.image("../../out/RData/2_Atlas_Var_Calc.RData")
```