---
title: "Merging mutliple Atlases"
author: "Friederike Wölke"
format: html
editor: visual
---

# 2) Handling multiple datasets

```{r}
#| warning: false
#| message: false
rm(list=ls())
library(sf)
library(dplyr)
library(tibble)
sf::sf_use_s2(FALSE)
```

## 2.1) Variables

```{r}
# folder path to grid data
source_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Atlas best practices/Frescalo_test_data/" # We don't have the data here yet
grid_path <-  paste0(source_path, list.files(source_path, pattern="*.gpkg", all.files =  T, recursive = T))

# Folder where to save the output (if needed)
out_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/GitHub/BEAST_General_Procedures/Frescalo/Frescalo_out/" # change to your output folder

# Add third Atlas name to vector when it's ready:
datasets <- c("Bird_Atlas_Czechia", "Bird_Atlas_NY")
```

## 2.2) Geopackage

```{r}
#| warning: false
#| message: false
#| label: geopackage  

grid_list <- sapply(grid_path, st_read, simplify = F, USE.NAMES = T)
names(grid_list) <- datasets
grid_list$Bird_Atlas_NY$dataset <- "Birds_Atlas_NY"
```

**some Checks & Notes:**

-   some columns missing

-   NY and CZ grids have different CRS

```{r}
# in CZ atlas but not in NY
setdiff(names(grid_list$Bird_Atlas_Czechia), names(grid_list$Bird_Atlas_NY)) 

# in NY atlas but not in CZ
setdiff(names(grid_list$Bird_Atlas_NY), names(grid_list$Bird_Atlas_Czechia)) 

st_crs(grid_list[[1]]) == st_crs(grid_list[[2]])
# They have different CRS
```

### Add time period column (with 1,2 or 3 indicating the time period)

```{r}
## add column for time period (tp)
for (i in 1:length(grid_list)){
  x <- grid_list[[i]]
  start_times <- sort(unique(x$start_year))
  end_times <- sort(unique(x$end_year))
  time_periods <- data.frame(start_year = start_times,
                           end_year = end_times, 
                           tp = 1:length(end_times))
  grid_list[[i]] <- merge(x, time_periods, by=c("start_year", "end_year"), all.x=T)
}

```

## Example code to run Frescalo over multiple datasets at once:

(based on requirements for: https://github.com/FrsLry/frescalo/blob/main/scripts/all_cz_atlas.R)

I modified the code from the link above a bit so that it works without the specific information from each datasets (e.g. start years)

**Workflow:**

-   Enclose everything into a loop that loops through grid_list

-   start with reducing grid_list to the "geopackage" object specific for each dataset.

-   Start a second loop nested within the first to loop through the different time periods

-   Then run frescalo over each dataset and time period separatly.

```{r}
library(lubridate)
library(sparta)

# Enclose everything into a loop that loops through grid_list and start with reducing grid_list to the "geopackage" object specific for each dataset. Then run frescalo over each dataset separatly.

# Vector of CRS for each dataset (modify CRS in vector for each dataset !!)
CRS_v <- c(5514, 5514) # change CRS for second atlas (NY) and add CRS for third atlas (NZ) here

# testing loop:
i <- 1
y <- 1
##

for (i in 1:length(grid_list)) {
  
  # Grid file
  geopackage <- grid_list[[i]] %>%
    st_transform(., CRS_v[i]) %>%
    mutate(time_period = years(start_year))


  for (y in 1:length(unique(geopackage$tp))) {
    # ecological data
    myData <- geopackage %>%
      select(cellID, verbatim_name, tp, start_year, time_period) %>%
      filter(tp == y) %>%
      rename(site = cellID, taxa = verbatim_name, year = start_year) %>%
      st_drop_geometry()

    ## Get the euclidean distance between the grid cells
    # First extract the coordinates
coords <-
geopackage %>%
  filter(tp == y) %>%
  distinct(cellID, .keep_all = T) %>%
  st_coordinates() %>%
  cbind(
    geopackage %>%
      st_drop_geometry() %>%
      filter(tp == y) %>%
      distinct(cellID)
  )

## Create the distance matrix
dist <- as.matrix(dist(coords))
# dist[upper.tri(dist)] <- NA

## Transform into a df and names rows and cols 

### This throws an error for me because of duplicate row names....[FW, 11.12.2023]
dist <-
  as.matrix(dist) %>%
  as.data.frame() %>%
  mutate(cellID = coords$cellID) %>%
  remove_rownames() %>%
  column_to_rownames("cellID")

colnames(dist) <- coords$cellID

## Transform into a long format
myDistances <-
  dist %>%
  rownames_to_column(var = "x") %>%
  pivot_longer(cols = -x,
               names_to = "y",
               values_to = "dist") %>%
  as.data.frame()



    ## Null habitat data
    myHabitatData <-
      data.frame(site = coords$cellID, x = 1)

    ## Compute the weights
    myWeights <- createWeights(
      distances = myDistances,
      attributes = myHabitatData,
      dist_sub = 200,
      sim_sub = 100
    )


    ## Run Frescalo
    frescalo_results <- frescalo(
      Data = myData,
      frespath = paste0(getwd(), "/Frescalo_3a_windows.exe"),
      time_periods = myTimePeriods,
      site_col = "site",
      sp_col = "taxa",
      year = "year",
      Fres_weights = myWeights,
      sinkdir = out_path,
      phi = NULL
    )
  }


  ## Get sampling effort
  effort_atlas <- geopackage %>%
    filter(tp == y) %>%
    select(cellID, effort) %>%
    filter(effort != 0) %>%
    rename(atlasCards = effort) %>%
    distinct()

  ## Compute effort from frescalo output
  frescalo_results$stat$effortFresc <- 1 / frescalo_results$stat$Alpha

  ## Plot effort atlas vs frescalo
  pdf(paste0(out_path, "Figs/fresc_vs_cards_", unique(grid_list[[i]]$dataset),"_", y,".pdf"),
    height = 4.13, width = 5.83
  )

  frescalo_results$stat %>%
    select(Location, effortFresc) %>%
    rename(cellID = Location) %>%
    right_join(effort_atlas, by = "cellID") %>%
    ggplot(aes(effortFresc, atlasCards)) +
    geom_point() +
    geom_smooth() +
    theme_bw()

  dev.off()

  ## Maps
  
 
  
  pdf(paste0(out_path, "Figs/map_fresc_", unique(grid_list[[i]]$dataset),"_", y,".pdf"),
    width = 6.09, height = 3.46
  )

  frescalo_results$stat %>%
    select(Location, effortFresc) %>%
    rename(cellID = Location) %>%
    right_join(effort_atlas, by = "cellID") %>%
    # pivot_longer(cols = c("effortFresc", "atlasCards"),
    #              names_to = "effort_type",
    #              values_to = "effort") %>%
    ggplot() +
    geom_sf(aes(color = effortFresc, geometry = geom), size = 4) +
    scale_color_viridis_c() +
    # facet_wrap(~effort_type)+
    theme_bw()

  dev.off()


  pdf(paste0(out_path, "Figs/map_cardNumber_", unique(grid_list[[i]]$dataset),"_", y,".pdf"),
    width = 6.09, height = 3.46
  )

  frescalo_results$stat %>%
    select(Location, effortFresc) %>%
    rename(cellID = Location) %>%
    right_join(effort_atlas, by = "cellID") %>%
    # pivot_longer(cols = c("effortFresc", "atlasCards"),
    #              names_to = "effort_type",
    #              values_to = "effort") %>%
    ggplot() +
    geom_sf(aes(color = atlasCards, geometry = geom), size = 4) +
    scale_color_viridis_c() +
    # facet_wrap(~effort_type)+
    theme_bw()

  dev.off()
}

```
