---
title: "0_2_Atlas_Predictors_prep"
author: "Friederike Wölke"
format: html
editor: visual
---

# Predictors from the Atlas data

Note: this is the first script used to study whether temporal change can be predicted from patterns that are collected once and do not change through time that we can find in nature (i.e., hereafter referred to as 'static pattern').\
Here we will prepare the data that will be used to predict temporal change.

In theory, it should be possible to partition the variance in species-level temporal change data into several factors that might play a role in the prediction. These 'predictors' are classified into several groups that describe different characteristics of the species, their ecological niche, their evolutionary history, the 'arena' which is being studied, and geometrical/statistical forces that could determine how the model predicts change.

The study's novelty derives from including geometrical, statistical and non-biological drivers in predicting temporal biodiversity change. The idea is that the statistical, geometrical, spatial and temporal limits of our study area and data collection might determine partially the result that can be predicted from the data.

We would expect that if a larger extent or a different scale is studied, temporal change patterns might depend on the situation. For example, in a small country, the magnitude of population increase is limited by the countries' borders, although the species range is most probably not (if there is no barrier or human intervention - e.g. Hunting). Thus, larger magnitudes of change can be detected on cross-border studies than it is structurally possible on a national scale.\
\
In a similar manner, we would expect that the geometrical properties of our study area can determine the results that our models are able to predict. One example would be the distribution of the species range inside the country. Where the spread of a species across the 'arena' (i.e., study area) naturally limits the species range inside the arena by arbitrary (country) borders which might not match with ecological borders. In addition, solely due to geometrical constraints, more species can move northward with climate in a country that follows an elongated East-West direction compared to one that is North-South elongated. We suggest that the borders of the study area introduce a 'veil' to the actual underlying temporal change processes that are happening, thereby masking the true temporal change dynamics. However, since species assessments take place on a country scale, it may be possible to use the predictive strength of geometry and characteristics of the arena to enhance our predictive capacity on national scales.

#### Part 1: Predicting future change: can data from the 80s predict the temporal change that took place between the 80s and 00s? Part 2: Predicting past change: can data from the 00s predict the temporal change that took place between the 80s and 00s?

## Libraries

```{r, message = F}
rm(list=ls())
gc()

# Essential data wrangling packages
library(dplyr); # Data wrangling
library(rstatix); # Reorder levels function
library(tidyr);
library(tibble);
library(readxl);

# Plotting: 
library(ggplot2); # Plotting
library(ggthemes); # Map theme
library(cowplot);  # Gridding plots together
library(gridExtra); # Gridding plots together and printing for pdf 

# Phylogenetic stuff:
library(ape);

# Spatial stuff:
library(sf); # Spatial stuff 1
library(terra); # Spatial stuff 2 
library(rnaturalearth)

# SAC:
library(spdep);
library(sfdep);

```

## Data paths

```{r}
### The paths ===
source_atlas <- c("c:/Users/wolke/OneDrive - CZU v Praze/Datasets/Processed/Atlases/Replicated/")
source_predictors <- c("c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/PhD_Projects/StaticPredictors/Data/")
source_Git <- c("c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/GitHub/BEAST_General_Procedures/Project_Frieda/StaticPredictors/")

# folder path to output folder
out_path <- c(paste0(source_Git, "out/"))


# folder path to atlas data
source_paths <- c(paste0(source_atlas, "Birds_Atlas_Czechia/"), 
                  paste0(source_atlas, "Birds_Atlas_New_York/"), 
                  paste0(source_atlas, "Birds_atlas_Japan/"), 
                  paste0(source_atlas, "Birds_atlas_EBBA/"))

# create path to read in data and grids from variables
data_paths <- c(paste0(source_paths[1],"Birds_Atlas_Czechia_beast_data.rds"), 
                paste0(source_paths[2], "Birds_Atlas_New_York_beast_data.rds"), 
                paste0(source_paths[3], "Birds_atlas_Japan_beast_data.rds"),
                paste0(source_paths[4], "Birds_atlas_EBBA_beast_data.rds"))

grid_paths <- c(paste0(source_paths[1],"Birds_Atlas_Czechia_grid.gpkg"), 
                paste0(source_paths[2], "Birds_Atlas_New_York_grid.gpkg"), 
                paste0(source_paths[3], "Birds_atlas_Japan_grid.gpkg"),
                paste0(source_paths[4], "Birds_atlas_EBBA_grid.gpkg"))

atlas_names <- c("Birds_Atlas_Czechia", "Birds_Atlas_New_York","Birds_atlas_Japan", "Birds_atlas_EBBA")

# Define the desired order of factor levels
desired_levels <- factor(c("1", "2","4", "8", "16", "32", "64", "128"), ordered = T,  
                         levels = c("1", "2","4", "8", "16", "32", "64", "128")) 


```

```{r}
grid_paths # list of atlas grids
data_paths # list of beast occ data

big_tab_path <- paste0(out_path, "Big_table_CZ_JP_NY.csv") # Created in 0_1_Atlas_prep.qmd
tree_path <- paste0(source_predictors, "Weeks_et_at_2022/singe_bird_phylo.tre")
traits_path <- paste0(source_predictors, "AVONET/ELEData/TraitData/AVONET3_BirdTree.xlsx")
sac_path <- paste0(out_path, "SAC_data.csv") # Created in this script (code runs a long time ~ 2h, so here is the object)
AVONET_path <- paste0(source_predictors, "AVONET/ELEData/TraitData/AVONET3_BirdTree.xlsx")


############## Read Processed Data from 0_1_Atlas_prep.qmd script:

pres_dat_final <- readRDS(paste0(out_path, "pres_dat_final")) # species data per cell  
presence_sf_list <- readRDS(paste0(out_path, "presence_sf_list.rds"))
big_tab <- read.csv(file=paste0(out_path, "Big_table_CZ_JP_NY.csv")) %>% 
  reorder_levels(cell_grouping, desired_levels)
avonet <- read_excel(AVONET_path, sheet = 2)

```

## The Data

```{r, read and process data}
## We need the highest resolution data = cell_grouping = 1

# grid data
grids <- list()
for (a in seq_along(grid_paths)) {
    grids_a <-sapply("cell1grid", function(i) {
      st_read(grid_paths[[a]], paste(i), quiet = TRUE)  %>% 
        st_transform(crs = 4326) %>% 
        reorder_levels( cell_grouping, order=desired_levels)
      }, simplify = FALSE)
    grids[[a]] <- grids_a$cell1grid
}

# Raw species occurrence data
# Species data  =====================
presence_data_all <- list()
for (i in seq_along(data_paths)){
  pres_dat <- readRDS(data_paths[i])
  sy <- sort(unique(pres_dat$start_year)) #sy = start_year
  
## Add time-period column 
  pres_dat2 <- pres_dat %>%     
    filter(cell_grouping == 1) %>%
    mutate(tp = ifelse(start_year == sy[1], 1, 2)) %>% 
    reorder_levels( cell_grouping, order=desired_levels)
  
  ## Filter for cells that have been sampled twice ============== (!!)
  pres_dat3 <- pres_dat2 %>% ungroup() %>% 
    group_by(cell_label) %>%
    mutate(num_periods = n_distinct(tp)) %>%
    mutate(repeated = ifelse(num_periods == 2, 1, 0)) %>%
    filter(repeated == 1)
  
  ## Filter for species that have change data (sp that occur in both time periods)
  common_species <- pres_dat3 %>%
    group_by(verbatim_name) %>%
    summarise(num_periods = n_distinct(tp)) %>%
    filter(num_periods == 2) %>%
    pull(verbatim_name)
  
  pres_dat4 <- pres_dat3 %>% filter(verbatim_name %in% common_species)

  presence_data_all[[i]] <- pres_dat4
}

# remove tp3 from CZ data for easier handling of all datasets

## Create a list of all raw occurrance tables
presence_data_all[[1]] <- presence_data_all[[1]] %>% 
  filter(start_year != 2014)



```

# Predictors

```{r, atlas predictors}
# Extent
# Shape
# Center of Gravity (of all species together) (COG)
# Total Species Richness (Gamma)
# Grain
# Start & End Years
# Sampling period length (End years - Start years)
# Fractal dimension of the landscape !! That would be cool !!

tp_list <- list()
for (i in seq_along(presence_data_all)){
  
  time_periods_df <- presence_data_all[[i]] %>% ungroup() %>%
    select(dataset, start_year, end_year, tp) %>%
    mutate(SampPeriodLength = end_year-start_year) %>%
    mutate(years = case_when(
      start_year %in% c(1972) ~ "70s",
      start_year %in% c(1980, 1985) ~ "80s",
      start_year %in% c(1997) ~ "90s",
      start_year %in% c(2000, 2001) ~ "00s",
      start_year %in% c(2013, 2016) ~ "10s")) %>% 
    unique()
  tp_list[[i]] <- time_periods_df
}

tp_df <- plyr::rbind.fill(tp_list)

atlas_predictors_df1 <- big_tab %>% 
  select(dataset, Total_area, tp) %>% 
  distinct() %>%
  mutate(
    grain = case_when(
      dataset == atlas_names[1] ~ "10",
      dataset == atlas_names[2] ~ "5",
      dataset == atlas_names[3] ~ "20", 
      dataset == atlas_names[4] ~ "50"))

atlas_predictors_df2 <- left_join(atlas_predictors_df1, tp_df)


# CZ = 1985-1989; 2001-2003
# NYS = 1980-1985; 2000-2004
# JP = 1997-2002; 2016-2021
# EU = 1972-1995; 2013-2017



```

```{r, species predictors}

# Log Ratio
# Start & End Years
# Telfer
# AOO
# Relative Occupancy
# Slope of OAR
# D
# Co-Occurrence
# Autocorrelation of the distribution (thanks Carmen!)
# Mean Diversity Measures per species: mean(cell_richness), Total atlas SR / mean(cell_richness)
# Geometric attributes of the distribution (thanks Gabriel!)
# HWI
# Phylogenetic Distinctness
# Family, Order, Genus
# Climatic Niche
# Mean Land use type
# Life Strategy
# Insectivorous ? 
# Threat status: National, Global ?
# Global Range Size
# Migration Status
# Human Association ? 
# NDVI ? 
# Occurrence outside of the arena ?
# Topology/Elevation ?



species_data_bigtable <- big_tab

# Make dataframe with predictors ======================
names(species_data_bigtable)

predictors_df <- species_data_bigtable %>% 
  filter(cell_grouping == 1) %>%
  select(
    verbatim_name, dataset, tp, # Grouping variables
    log_R2_1, Telfer_1_2, # Response variables
    Total_area, # Extent of the Arena
    relative_occupancy_Ncells, # Relative Occupancy
    m_AOO_s, # slope of OAR1
    m_relOcc_s, # slope of OAR2
    D_AOO_s, # D of AOO
    D_relOcc_s # D of rel. Occupancy
    )

```

## Diversity Metrics

```{r, eval = F}
## Calculate Diversity Measures ======================
GammaAlphaBeta_Atlas <- pres_dat_final %>%
  filter(cell_grouping == 1) %>%
  select(dataset, tp, cell_label, verbatim_name) %>% distinct() %>%
  group_by(dataset,tp) %>%
  mutate(GammaSR = sum(n_distinct(verbatim_name))) %>% ungroup() %>%
  group_by(dataset, tp, cell_label) %>%
  mutate(AlphaSR = sum(n_distinct(verbatim_name))) %>%
  mutate(BetaSR = GammaSR/AlphaSR)

GammaAlphaBeta_Species <- GammaAlphaBeta_Atlas %>%   
  ungroup() %>%
  group_by(dataset, tp, verbatim_name) %>%
  mutate(AlphaSR_sp = mean(AlphaSR)) %>%
  mutate(BetaSR_sp = GammaSR/AlphaSR_sp) %>%
  select(dataset, tp, verbatim_name, AlphaSR_sp, BetaSR_sp, GammaSR) %>% distinct()

# predictors_df2 <- merge(predictors_df, GammaAlphaBeta_Species, by=c("dataset", "tp", "verbatim_name"), all = T)

## Calculate Mean Sampling Effort  ======================
AvgEffort <- pres_dat_final %>%
  filter(cell_grouping == 1) %>%
  select(dataset, tp, cell_label, verbatim_name, samp_effort_type, effort) %>% 
  distinct() %>%
  group_by(dataset, tp, verbatim_name) %>%
  summarize(avgEffort = mean(effort))


Diversity_Effort <- full_join(GammaAlphaBeta_Species, AvgEffort)
saveRDS(Diversity_Effort, paste0(out_path, "Diversity_AvgEffort.rds"))
```

#### Diversity_Effort: Read back in
```{r}

Diversity_Effort <- readRDS(paste0(out_path, "Diversity_AvgEffort.rds"))
predictors_df2 <- full_join(predictors_df, Diversity_Effort)

```

## Spatial Autocorrelation

ref: Carmen

```{r, eval = F}
## autocorrelation loop

sf_use_s2(FALSE)
out_list <- list()
time_periods <- c(1,2)
atlas_names <- c("Birds_Atlas_Czechia", "Birds_Atlas_New_York","Birds_atlas_Japan", "Birds_atlas_EBBA")

```

This takes quite some time (~ 2.5h). Read the SAC_df back in instead of re-running the code.
```{r, eval = F}

presence_sf_list <- list()
for (a in seq_along(atlas_names)){
  grid <- grids[[a]]
  presence_data <- presence_data_all[[a]]
  presence_sf <- left_join(grid, presence_data)
  dd <- presence_sf %>% filter(dataset == atlas_names[a])
  presence_sf_list[[a]] <- dd


  for (y in seq_along(time_periods)){
    dd1 <- dd %>% filter(tp == time_periods[y])
    
    ## From Carmen ===============================
    
    sp_list <- unique(dd1$verbatim_name)
    # Contiguity methods
    neighbor_method <- "contiguity"
    # Output columns
    columns_df <- c("tp", "dataset", "neighbor_method", "mean_area", "mean_area_cropped")
    spmor_cols <- c("verbatim_name", "global_moran", "p_value", columns_df)
    spmoran_df <- data.frame(matrix(nrow = length(sp_list), ncol = length(spmor_cols)))
    colnames(spmoran_df) <- spmor_cols

    row_id <- 1
    
        
    for (s in seq_along(sp_list)) {
      
      print(paste("atlas =", atlas_names[a], "tp =", time_periods[y], "sp = ", sp_list[s]))
      # Obtain a df for each species that indicates the presence (1) or absence (0) of the species
      data_species <- dd1 %>%
        mutate(verbatim_name = ifelse(verbatim_name != sp_list[s], NA, verbatim_name)) %>%
        group_by(across(-verbatim_name)) %>%
        slice(which.max(!is.na(verbatim_name))) %>%
        mutate(presence = ifelse(!is.na(verbatim_name), 1, 0))
      
      # Filling in the data
      spmoran_df$verbatim_name[row_id] <- sp_list[s]
      spmoran_df$dataset[row_id] <- atlas_names[a]
      spmoran_df$neighbor_method[row_id] <- neighbor_method
      spmoran_df$tp[row_id] <- unique(data_species$tp)
      spmoran_df$mean_area[row_id] <- mean(data_species$area, na.rm = TRUE)
      spmoran_df$mean_area_cropped[row_id] <- mean(data_species$area_cropped, na.rm = TRUE)

      tryCatch({
      # Creating neighbors
      nb <- poly2nb(data_species, queen = TRUE)
      # Spatial weights for the neighbors
      listw <- nb2listw(nb, style = "W", zero.policy = TRUE)
      # Moran's I
      moran_res <- moran.test(data_species$presence, listw, zero.policy = TRUE)
      spmoran_df[row_id, "global_moran"] <- as.numeric(moran_res$estimate)[1]
      spmoran_df[row_id, "p_value"] <- moran_res$p.value
      }, error = function(e) {
        # Handle the case where Moran's I cannot be calculated
        spmoran_df[row_id, c("global_moran", "p_value")] <- NA
      })

      row_id <- row_id + 1  # Update row_id for the next entry
    
      }
    
    out_list[[length(out_list) + 1]] <- spmoran_df
  
  }
  
}
saveRDS(presence_sf_list, paste0(out_path, "presence_sf_list.rds"))
SAC_df <- plyr::rbind.fill(out_list, fill = T) %>% unique()
write.csv(SAC_df, paste0(out_path, "SAC_data.csv"))
saveRDS(SAC_df, paste0(out_path, "SAC_data.rds"))

```

#### SAC_df: Read back in
```{r}
SAC_df <- readRDS(paste0(out_path, "SAC_data.rds"))

predictors_df4 <- full_join(predictors_df2, SAC_df)
predictors_df4

```

## Characterize Geometries

(Function by Dr. Gabriel Ortega: requires terra & tidyverse)

```{r, geometries function from Gabriel}
pacman::p_load(terra, tidyverse, tidyterra)
# Function to get the max elongation per polygon, north-south length or east-west length, circularity, etc...
poly_attr <- function(x, type = NULL) {
  # Convert "sf" to SpatVector
  if (inherits(x, "sf") == T) {
    x <- vect(x)
  }
  # Reproject to WGS84
  x <- project(x, "epsg:4326")
  # Get East-West terra::distance (longitudinal extension)
  if (type == "ew") {
    pol <- as.data.frame(crds(as.points(ext(x))))
    sw <- pol %>%
      summarise(x = min(x), y = min(y)) %>%
      as.matrix()
    se <- pol %>%
      summarise(x = max(x), y = min(y)) %>%
      as.matrix()
    res <- terra::distance(sw, se, lonlat = T)[[1]]
  }
  # Get South-North terra::distance (latitudinal extension)
  if (type == "ns") {
    pol <- as.data.frame(crds(as.points(ext(x))))
    sw <- pol %>%
      summarise(x = min(x), y = min(y)) %>%
      as.matrix()
    nw <- pol %>%
      summarise(x = min(x), y = max(y)) %>%
      as.matrix()
    res <- terra::distance(sw, nw, lonlat = T)[[1]]
  }
  # Get max terra::distance between oposite vertices
  if (type == "max") {
    pol <- crds(as.points(ext(x)))
    res <- terra::distance(pol, lonlat = T) %>% max()
    res <- res
  }
  # Circularity
  if (type == "circ") {
    perimeter <- perim(x)
    area <- expanse(x)
    res <- (perimeter^2) / area
  }
  # Normalized circularity
  if (type == "circN") {
    perimeter <- perim(x)
    area <- expanse(x)
    res <- (perimeter^2) / (4 * pi * area)
  }
  # Major length of the minimum rectangle
  if (type == "majElon") {
    minRectangle <- minRect(x)
    df <- as.vector(terra::distance(crds(as.points(minRectangle)), lonlat = T))
    df <- sort(df)
    res <- mean(df[[3]], df[[4]])
  }
  # Width of the minimum rectangle
  if (type == "majWidth") {
    minRectangle <- minRect(x)
    df <- as.vector(terra::distance(crds(as.points(minRectangle)), lonlat = T))
    df <- sort(df)
    res <- mean(df[[1]], df[[2]])
  }
  # Elongation ratio (I AM NOT TOTALLY SURE IF THIS FITS THE DEFINITION OF ELONGATION)
  if (type == "elon") {
    minRectangle <- minRect(x)
    df <- as.vector(terra::distance(crds(as.points(minRectangle)), lonlat = T))
    df <- sort(df)
    length <- mean(df[[3]], df[[4]])
    width <- mean(df[[1]], df[[2]])
    res <- 1 - (width / length)
  }
  res <- as.numeric(res)
  return(res)
}
```

### Calculate Geometries: 

This takes quite some time (~ 2 hours). Better read the objects back in below.
```{r, calculate geometries}

geom_ls <- list()
for (i in seq_along(grids)){
  
  test <- grids[[i]] %>% 
    select(geom, cell_label) %>%
    summarise() %>% 
    terra::vect()
  
  Geom_attributes <- data_frame(dataset = atlas_names[i],
           NS = poly_attr(test, "ns"),
           EW = poly_attr(test, "ew"),
           MAX = poly_attr(test, "max"),
           MajELON = poly_attr(test, "majElon"),
           MajWIDTH = poly_attr(test, "majWidth"),
           ELON = poly_attr(test, "elon"),
           CIRC = poly_attr(test, "circ"),
           CIRCN = poly_attr(test, "circN"))
  geom_ls[[i]] <- Geom_attributes
}
countries_geom_attributes <- plyr::rbind.fill(geom_ls) 


### For each species separately:
species_geom <- list()
year_species_geom <- list()
atlas_all_sp_geom <-list()
COG_all <- list()

for (a in seq_along(presence_sf_list)){
  atlas <- presence_sf_list[[a]]
  COG <- atlas %>% summarise(CenterOfGravity_Atlas_long = mean(cell_long),
                       CenterOfGravity_Atlas_lat = mean(cell_lat),
                       dataset = unique(dataset))
  COG_all[[a]] <- COG
  
  
  for (y in seq_along(time_periods)){
    atlas_y <- atlas %>% filter(tp == time_periods[y])
    sp <- unique(atlas_y$verbatim_name)
    
    for (s in seq_along(sp)){
      atlas_sp <- atlas_y %>% filter(verbatim_name == sp[s])
      ## Maybe I have to summarize here ! Not sure - let's try this and compare.
      
      test <- atlas_sp %>% 
        select(geom, cell_label)%>% 
        summarise() %>% 
        vect()
       
      ## Some plots
      minRectangle <- minRect(test)
      plet(minRectangle)

      minCircle <- minCircle(test)
      plet(minCircle)

      convexHull <- convHull(test)
      plet(convexHull)
      ####
 
 
      Geom_attributes <- data_frame(dataset = atlas_names[a],
                                    tp = time_periods[y],
                                    verbatim_name = sp[s],
                                    NS = poly_attr(test, "ns"),
                                    EW = poly_attr(test, "ew"),
                                    MAX = poly_attr(test, "max"),
                                    MajELON = poly_attr(test, "majElon"),
                                    MajWIDTH = poly_attr(test, "majWidth"),
                                    ELON = poly_attr(test, "elon"),
                                    CIRC = poly_attr(test, "circ"),
                                    CIRCN = poly_attr(test, "circN"))

  species_geom[[s]] <- Geom_attributes
  }
  species_geom_df <- plyr::rbind.fill(species_geom)  
  year_species_geom[[y]] <- species_geom_df  
  }
  year_species_geom_df <- plyr::rbind.fill(year_species_geom) 
  atlas_all_sp_geom[[a]] <- year_species_geom_df
}

Species_geom_attributes <- plyr::rbind.fill(atlas_all_sp_geom) 
Geometries <- full_join(Species_geom_attributes, countries_geom_attributes)
COG_df <- plyr::rbind.fill(COG_all)
Geometries <- full_join(Geometries, COG_df)

saveRDS(Geometries, paste0(out_path, "Geometries.rds"))
saveRDS(COG_all, paste0(out_path, "CenterOfGravity_list.rds"))

```
#### Geometries: Read back in
```{r}


Geometries <- readRDS(paste0(out_path, "Geometries.rds"))

predictors_df5 <- full_join(Geometries, predictors_df4) 
```


## Co-Occurrence

```{r}

species_data <- pres_dat_final %>% filter(cell_grouping == 1) %>% select(dataset, tp, verbatim_name, cell_label) %>% distinct()

# ----------------------------------------------------------- #
co_occ_list1<- list()
for (a in seq_along(atlas_names)){
  
  # create community dataframe
  species_data_a <- species_data %>% filter(dataset == atlas_names[a])
  comm_dat <- species_data_a %>% 
    group_by(tp) %>% 
    dplyr::select(tp, verbatim_name, cell_label) %>%
    ungroup() %>% 
    distinct()
  comm_dat <- as.data.frame(comm_dat)
  
  # Convert data frame to Species X sites matrix
  comm_matrix_tp1 <- fossil::create.matrix(comm_dat, 
                                           tax.name = "verbatim_name", 
                                           locality = "cell_label", 
                                           time.col = "tp", 
                                           time = "1", 
                                           abund = F)
  comm_matrix_tp2 <- fossil::create.matrix(comm_dat, 
                                           tax.name = "verbatim_name", 
                                           locality = "cell_label", 
                                           time.col = "tp", 
                                           time = "2", 
                                           abund = F)
  
  # Calculate probability of pairwise co-occurrence of species: ----------- #
  co_occurrence_tp1 <- cooccur::cooccur(comm_matrix_tp1, spp_names = T)
  print(length(unique(co_occurrence_tp1$spp.names))) # 206 species 
  co_occurrence_tp2 <- cooccur::cooccur(comm_matrix_tp2, spp_names = T)
  print(length(unique(co_occurrence_tp2$spp.names))) # 213 species
  
  co_occ_list1[[a]] <- list(co_occurrence_tp1, co_occurrence_tp2)
  
  
}

co_occ_list2 <-list()
for (a in seq_along(atlas_names)){
  species_data_a <- species_data %>% filter(dataset == atlas_names[a])
  species_data_a1 <- species_data_a %>% select(-cell_label) %>% distinct()

  res1 <- data.frame(co_occ_list1[[a]][[1]]$results) #205 species
  res1 <- res1 %>% group_by(sp1_name) %>%
    mutate(mean_prob_cooccur = mean(prob_cooccur), tp = "1") %>% 
    dplyr::select(sp1_name, mean_prob_cooccur, tp) %>% 
    distinct()

  res2 <- data.frame(co_occ_list1[[a]][[2]]$results) #212 species
  res2 <- res2 %>% group_by(sp1_name) %>%
    mutate(mean_prob_cooccur = mean(prob_cooccur), tp = "2") %>% 
    dplyr::select(sp1_name, mean_prob_cooccur, tp) %>% 
    distinct()
  
  results <- full_join(res1,res2)
  
  species_data_a1_2 <- merge(species_data_a1, results, by.x=c("verbatim_name", "tp"), by.y = c("sp1_name", "tp"), all = T)
  
  co_occ_list2[[a]] <- species_data_a1_2

}
 
Co_Occ_df_final <- plyr::rbind.fill(co_occ_list2)
saveRDS(Co_Occ_df_final, paste0(out_path, "Coocc_df_final.rds"))


# Some summaries -------------------------------------------------------- #

# summary(co_occurrence_tp1)
# prob.table(co_occurrence_tp1)
# plot(co_occurrence_tp1, cex.axis = 0.5)
# pair.profile(co_occurrence_tp1)
# obs.v.exp(co_occurrence_tp1)




```
#### Co_Occ_df_final: Read back in 
```{r}

Co_Occ_df_final <- readRDS(paste0(out_path, "Coocc_df_final.rds"))

predictors_df6 <- full_join(Co_Occ_df_final, predictors_df5) 
```

## AVONET Traits

```{r}
names(avonet)
avonet <- avonet %>% 
  select("Species3", "Family3", "Order3", "Hand-Wing.Index", "Mass", "Habitat", "Habitat.Density", "Migration", "Trophic.Level", "Trophic.Niche", "Primary.Lifestyle", "Range.Size") %>% 
  rename("verbatim_name"= "Species3") %>%
  mutate_at(vars("Range.Size"), as.numeric)

predictors_df7 <- left_join(predictors_df6, avonet) %>% distinct()

```

#### Merging Atlas + Species predictors

```{r}
all_predictors0 <- full_join(predictors_df7, atlas_predictors_df2)
str(all_predictors0)

saveRDS(all_predictors0, paste0(out_path, "All_predictors.rds"))
```

## Correlation Analysis
```{r}
all_predictors <- all_predictors0 %>%
  select(-neighbor_method, -verbatim_name) %>%
  mutate_at(
    c("dataset", "years", "tp", "Order3", "grain", "Family3", "Habitat", "Migration", "Trophic.Level", "Trophic.Niche", "Primary.Lifestyle"), as.factor
  ) %>% 
  st_drop_geometry() %>% select(-geom) %>% na.omit()

summary(lm(log_R2_1 ~ .,data = all_predictors, na.action = "na.omit"))


library(ggcorrplot)
cor_df <- all_predictors %>% select(-Order3, -grain, -years, -SampPeriodLength)

p.mat <-  model.matrix(~0+., data=cor_df) %>% ggcorrplot::cor_pmat()

model.matrix(~0+., data=cor_df) %>% 
  cor(use="pairwise.complete.obs") %>% ggcorrplot(tl.cex = 5, type = "lower",outline.color = "white")

  ggcorrplot( lab=TRUE, lab_size=2, hc.order=T, insig ="blank",p.mat = p.mat)

```
