---
title: "0_Data_preparation"
author: "Friederike Wölke"
format: html
editor: visual
---

# Predicting change from static patterns

In this project I want to look at different static patterns that describe the distribution of a species and could potentially used to predict temporal change

## Structure:

-   Data:

    -   Birds of CZ, Plants of CZ

    -   Birds of Europe, Plants of Europe

    -   IUCN global range maps for co-occurrence patterns, WCVP for Plant co-occurrences OR\
        BirdLife International range maps?

    -   AVONET trait data

        -   e.g. Dispersal: Hand-Wing-Index (HWI) (ref: flight-efficiacy: Weeks et al 2022: https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2435.14056)

    -   BirdTree (Jetz et al. 2012) Phylogeny to calculate a measure of phylogenetic distance/distinctiveness

## Set up libraries:

```{r}
#| message = F
library(sf)
library(dplyr)
library(ape)
library(readxl)
library(phyloregion)
library(randomForest)
library(randomForestExplainer)
library(vegan)
```

## Set up Variables:

-   source_path = path to Atlas

-   out_path = path to save output from R

-   data_path = filename of the data

-   grid_path = filename of the grid

-   tree_path = filename of phylogenetic tree to calculate phylo metrics

-   traits_path = filename of AVONET matched to BirdTree

```{r}
#| warning: false
#| message: false
#| label: Variables

rm(list=ls())
# folder path to atlas data
source_path_country <- "c:/Users/wolke/OneDrive - CZU v Praze/Datasets/Processed/Atlases/Replicated/Birds_Atlas_Czechia/"

# folder path to output folder
out_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/GitHub/StaticPredictors/StaticPredictors/StaticPredictors/out/"

# create path to read in data and grids from variables
data_path_country <- paste0(source_path_country,"Birds_Atlas_Czechia_beast_data.rds")
grid_path_country <-  paste0(source_path_country,"Birds_Atlas_Czechia_grid.gpkg")

# create paths to read phylogenetic tree and traits for birds (BirdTree, AVONET)
tree_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/R/Weeks_et_at_2022/singe_bird_phylo.tre"
traits_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/R/AVONET/ELEData/TraitData/AVONET3_BirdTree.xlsx"
fractal_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/GitHub/BEAST_General_Procedures/Project_Frieda/out/Big_table.csv"

```

### Read in data

# !!! FOR NOW - I will make a subset for 30 species because the data is too big

```{r}

# Species distributions  ---------------------------------------------------------------------- #
species_data_country <- readRDS(data_path_country)
species_data_country$cell_grouping <- as.factor(species_data_country$cell_grouping)

# Species Fractal Data ------------------------------------------------------------------------ #
fractal_df <- read.csv(fractal_path)
sp_list <- unique(fractal_df$verbatim_name)
# ----------------------------------------------------------------- ===   CREATE SUBSET HERE    # 

# Sample 30 species                                                          
set.seed(123)
subset <- sample(sp_list, 30)
species_data_subset <- species_data_country %>% filter(verbatim_name %in% subset)
fractal_df_subset <- fractal_df %>% filter(verbatim_name %in% subset) %>% select(-X)


# phylogenetic tree --------------------------------------------------------------------------- #
tree <- ladderize(read.tree(tree_path))

# bird traits --------------------------------------------------------------------------------- #
avonet <- read_excel(traits_path, sheet = 2)
avonet <- avonet %>% filter(Species3 %in% subset)

# grid data------------------------------------------------------------------------------------ #
layers_country <- st_layers(grid_path_country)$name

grid_list_country <- sapply(layers_country, function(i) {
  st_read(grid_path_country, paste(i), quiet = TRUE)
}, simplify = FALSE)


```

#### Manipulate the data / Calculate Phylogenetic Distinctness for each species (as fair proportion)

1.  create column in species data that matches the tip-label formating in the tree (here with "\_" between genus name and species epithet)
2.  Calculate evolutionary distinctness (fair proportion metric) (FD)
3.  extract Hand wing index (and range size?) from AVONET
4.  Rename and sort grid-sizes to match grid and data (from small to large grid)
5.  Create column for time periods (1-3 for CZ birds)
6.  Rename the spatial scale to a relative scale, where 1 indicates the largest grid size that covers the whole study 'arena' (in grid and data)
7.  Split the species data into time periods

##### Country Data

```{r}

# --------------------------------------------------------------------------------------------- #
# adjust species names in data to match those in the tree
species_data_subset$tip.label <- gsub(" ", "_", species_data_subset$verbatim_name)
# species_data_continent$tip.label <- gsub(" ", "_", species_data_continent$verbatim_name)

FP <- evol_distinct(tree, "fair.proportion") # add column with phylogenetic distinctness
FP <- data.frame(FP)

species_data_subset <- merge(species_data_subset, FP, by.x = "tip.label", by.y="row.names", all.x=T)

# --------------------------------------------------------------------------------------------- #

avonet$tip.label <- gsub(" ", "_", avonet$Species3)
avonet <- avonet %>% select(tip.label, Family3, Order3, "Hand-Wing.Index", "Range.Size")

species_data_subset <- merge(species_data_subset, avonet, by = "tip.label", all.x=T)

# --------------------------------------------------------------------------------------------- #

# save names of layers in data (needed because now they don't match anymore between grid and data)
layers_country <- unique(species_data_subset$cell_grouping)


# Define the desired order of factor levels
## Change cell_grouping here (!) after Gabriel has modified it                            [# !!!!!]
desired_levels_data_country <- factor(layers_country, 
                         ordered = T, 
                         levels = c("1", "2", 
                                    "4", "8", 
                                    "16", "32", 
                                    "64")) 

# --------------------------------------------------------------------------------------------- #

## sort the cell groupings ascending
species_data_subset <- species_data_subset %>%
  mutate(cell_grouping = factor(cell_grouping, 
                                levels = desired_levels_data_country))

# --------------------------------------------------------------------------------------------- #

## add column for time period (tp) - Country
start_times <- sort(unique(species_data_subset$start_year))
end_times <- sort(unique(species_data_subset$end_year))

time_periods <- data.frame(start_year = start_times,
                           end_year = end_times, 
                           tp = seq_along(end_times))

species_data_subset <- merge(species_data_subset, time_periods, 
                       by=c("start_year", "end_year"), 
                       all.x=T)

# --------------------------------------------------------------------------------------------- #

# create scale column as a fraction of the full country:
species_data_subset <- species_data_subset %>% mutate(
      scale = case_when(
        cell_grouping == as.character(desired_levels_data_country[1]) ~ 1/64,
        cell_grouping == as.character(desired_levels_data_country[2]) ~ 1/32,
        cell_grouping == as.character(desired_levels_data_country[3]) ~ 1/16,
        cell_grouping == as.character(desired_levels_data_country[4]) ~ 1/8,
        cell_grouping == as.character(desired_levels_data_country[5]) ~ 1/4 ,
        cell_grouping == as.character(desired_levels_data_country[6]) ~ 1/2,
        cell_grouping == as.character(desired_levels_data_country[7]) ~ 1)) %>% 
  unique()

# --------------------------------------------------------------------------------------------- #

species_data_subset_merged <- merge(species_data_subset, fractal_df_subset, by = c("verbatim_name", "scale", "tp", "cell_grouping"))
species_data_subset_merged <- unique(species_data_subset_merged)

cell_columns <- c("area", "cell_label", "area_cropped", "cell_long", "cell_lat", "effort", "samp_effort_type")
species_data_subset_merged_short <- species_data_subset_merged %>% select(!cell_columns) %>% distinct()


# --------------------------------------------------------------------------------------------- #



sp_dat1 <- species_data_subset_merged %>% filter(tp == 1 & cell_grouping == "1")
sp_dat2 <- species_data_subset_merged %>% filter(tp == 2 & cell_grouping == "1")
sp_dat3 <- species_data_subset_merged %>% filter(tp == 3 & cell_grouping == "1")

# --------------------------------------------------------------------------------------------- #
# geopackage grids:
layers_country <- st_layers(grid_path_country)$name

desired_levels_grid_country <- factor(layers_country, 
                         ordered = T, 
                         levels = c("cell1grid", "cell2grid", 
                                    "cell4grid", "cell8grid", 
                                    "cell16grid", "cell32grid", 
                                    "cell64grid")) 

# grid data
grid_list_country <- sapply(layers_country, function(i) {
  st_read(grid_path_country, paste(i), quiet = TRUE)
}, simplify = FALSE)

# add scale column to the grid
for (i in seq_along(grid_list_country)) {
  grid_list_country[[i]] <- grid_list_country[[i]] %>% mutate(
      scale = case_when(
        cell_grouping == as.character(desired_levels_grid_country[1]) ~ 1/64,
        cell_grouping == as.character(desired_levels_grid_country[2]) ~ 1/32,
        cell_grouping == as.character(desired_levels_grid_country[3]) ~ 1/16,
        cell_grouping == as.character(desired_levels_grid_country[4]) ~ 1/8,
        cell_grouping == as.character(desired_levels_grid_country[5]) ~ 1/4 ,
        cell_grouping == as.character(desired_levels_grid_country[6]) ~ 1/2,
        cell_grouping == as.character(desired_levels_grid_country[7]) ~ 1)) %>% 
  unique()
}

# --------------------------------------------------------------------------------------------- #

communitymatrix(species_data_country)

```

##### Continental Data

```{r, Continent data}

# species_data_continent <- readRDS(data_path_continent)
# layers_continent <- st_layers(grid_path_continent)$name

# grid_list_continent <- sapply(layers_continent, function(i) {
#   st_read(grid_path_continent, paste(i), quiet = TRUE)
# }, simplify = FALSE)
# species_data_continent <- merge(species_data_continent, FP, by.x = "tip.label", by.y="row.names", all.x=T)
# species_data_continent <- merge(species_data_continent, avonet, by = "tip.label", all.x=T)
# layers_continent <- unique(species_data_continent$cell_grouping)
# desired_levels_data_continent <- factor(layers_continent, 
#                          ordered = T, 
#                          levels = c("cell1grid", "cell2grid", 
#                                     "cell4grid", "cell8grid", 
#                                     "cell16grid", "cell32grid", 
#                                     "cell64grid", "cellfullgrid")) 
# 
# species_data_continent <- species_data_continent %>%
#   mutate(cell_grouping = factor(cell_grouping, 
#                                 levels = desired_levels_data_continent))
# ## add column for time period (tp) - Continent
# start_times <- sort(unique(species_data_continent$start_year))
# end_times <- sort(unique(species_data_continent$end_year))
# 
# time_periods <- data.frame(start_year = start_times,
#                            end_year = end_times, 
#                            tp = seq_along(end_times))
# 
# species_data_continent <- merge(species_data_continent, time_periods, 
#                        by=c("start_year", "end_year"), 
#                        all.x=T)
# species_data_continent <- species_data_continent %>% mutate(
#       scale = case_when(
#         cell_grouping == as.character(desired_levels_data[1]) ~ 1/128,        
#         cell_grouping == as.character(desired_levels_data[2]) ~ 1/64,
#         cell_grouping == as.character(desired_levels_data[3]) ~ 1/32,
#         cell_grouping == as.character(desired_levels_data[4]) ~ 1/16,
#         cell_grouping == as.character(desired_levels_data[5]) ~ 1/8,
#         cell_grouping == as.character(desired_levels_data[6]) ~ 1/4 ,
#         cell_grouping == as.character(desired_levels_data[7]) ~ 1/2,
#         cell_grouping == as.character(desired_levels_data[8]) ~ 1)) %>% 
#   unique()

# # CONTINENT
# 
# layers_continent <- st_layers(grid_path_continent)$name
# 
# desired_levels_grid_continent <- factor(layers, 
#                          ordered = T, 
#                          levels = c("cell1grid", "cell2grid", 
#                                     "cell4grid", "cell8grid", 
#                                     "cell16grid", "cell32grid", 
#                                     "cell64grid", "cell128grid")) 
# 
# # grid data
# grid_list_continent <- sapply(layers_continent, function(i) {
#   st_read(grid_path_continent, paste(i), quiet = TRUE)
# }, simplify = FALSE)
# 
# # add scale column to the grid
# for (i in seq_along(grid_list_continent)) {
#   grid_list_continent[[i]] <- grid_list_continent[[i]] %>% mutate(
#       scale = case_when(
#         cell_grouping == as.character(desired_levels_grid_continent[1]) ~ 1/64,
#         cell_grouping == as.character(desired_levels_grid_continent[2]) ~ 1/32,
#         cell_grouping == as.character(desired_levels_grid_continent[3]) ~ 1/16,
#         cell_grouping == as.character(desired_levels_grid_continent[4]) ~ 1/8,
#         cell_grouping == as.character(desired_levels_grid_continent[5]) ~ 1/4 ,
#         cell_grouping == as.character(desired_levels_grid_continent[6]) ~ 1/2,
#         cell_grouping == as.character(desired_levels_grid_continent[7]) ~ 1)) %>% 
#   unique()
# }

```

##### Drop Species from Tree ?

```{r, drop tips from tree}
# # for each time period separately:
# # a) extract species lists for each time period
# sp_list_df1 <- sp_dat1  %>% distinct(tip.label)
# sp_list_df2 <- sp_dat2  %>% distinct(tip.label)
# sp_list_df3 <- sp_dat3  %>% distinct(tip.label)
# 
# # b) create list of species not in the data (to drop them from the tree)
# sp.drop.tree1 <- setdiff(tree$tip.label, sp_list_df1$tip.label)
# sp.drop.tree2 <- setdiff(tree$tip.label, sp_list_df2$tip.label)
# sp.drop.tree3 <- setdiff(tree$tip.label, sp_list_df3$tip.label)
# 
# # c) create list of species not in the tree
# sp.drop.data1 <- setdiff(sp_list_df1$tip.label, tree$tip.label) 
# sp.drop.data2 <- setdiff(sp_list_df2$tip.label, tree$tip.label) 
# sp.drop.data3 <- setdiff(sp_list_df3$tip.label, tree$tip.label) 
# 
# 
# # --------------------------------------------------------------------------------------------- #
# # drop species from tree (not in data):
# # -- time period 1 -- #
# tree1 <- drop.tip(tree, sp.drop.tree1)
# tree1 <- ladderize(tree1)
# # -- time period 2 -- #
# tree2 <- drop.tip(tree, sp.drop.tree2)
# tree2 <- ladderize(tree2)
# # -- time period 3 -- #
# tree3 <- drop.tip(tree, sp.drop.tree3)
# tree3 <- ladderize(tree3)
# 
# 
# # --------------------------------------------------------------------------------------------- #
# # drop species from data (not in tree)
# # -- time period 1 -- #
# sp_dat1 <- sp_dat1 %>% filter(verbatim_name %in% sp.drop.data1)
# # -- time period 2 -- #
# sp_dat2 <- sp_dat2 %>% filter(verbatim_name %in% sp.drop.data2)
# # -- time period 3 -- #
# sp_dat3 <- sp_dat3 %>% filter(verbatim_name %in% sp.drop.data3)

```

```{r}
model_df <- species_data_subset_merged_short
model_df <- model_df %>% rename("HWI" = "Hand-Wing.Index")

str(model_df)
model_df$tp <- as.factor(model_df$tp)
model_df$cell_grouping <- factor(model_df$cell_grouping, levels = c("1", "2", "4", "8", "16", "32", "64"), labels = c("1", "2", "4", "8", "16", "32", "64"))
model_df$start_year <- as.factor(model_df$start_year)
model_df$end_year <- as.factor(model_df$end_year)

model_df$dataset <- as.factor(model_df$dataset)
model_df$taxa <- as.factor(model_df$taxa)
model_df$Family3 <- as.factor(model_df$Family3)
model_df$Order3 <- as.factor(model_df$Order3)

names(model_df)
model_df_ss <- model_df %>% 
  select(-cell_grouping, -start_year, -end_year, -tip.label, -dataset, -license, -taxa )


# --------------------------------------------------------------------------------- #
# install.packages("randomForestExplainer")
library(randomForestExplainer)

# Data partitioning =====
set.seed(123)

samp <- sample(nrow(model_df_ss), 0.8 * nrow(model_df_ss))
train <- model_df_ss[samp, ]; dim(train)
test <- model_df_ss[-samp, ]; dim(test)

# Model ====
sp.rf <- randomForest(r2_AOO ~ ., data=train, ntree=1000, mtry=5, importance=TRUE, proximity=TRUE, na.action = na.omit,localImp = TRUE) # explains 90% of variance!! WOW?!??!

sp.rf$coefs #? empty?
round(sp.rf$importance,3)

# Variable importance =====
round(importance(sp.rf),2)
varImpPlot(sp.rf)
impt_frame<-measure_importance(sp.rf)
plot_multi_way_importance(impt_frame, no_of_labels = 6)
plot_multi_way_importance(impt_frame, x_measure = "accuracy_decrease", y_measure = "gini_decrease", size_measure = "p_value", no_of_labels = 6)


prediction <- predict(sp.rf, newdata = test)

table(prediction, test$r2_AOO)

prediction

min_depth_frame <- min_depth_distribution(sp.rf)
plot_min_depth_distribution(min_depth_frame)
# ----------------------------------- #
install.packages("ranger")
library(ranger)

rf2 <- ranger(r2_AOO ~ ., data=train, importance = "permutation", tuneGrid = data.frame(mtry = ceiling((ncol(model_df_ss)-1)/3), splitrule = "variance", min.node.size = 5))
pred <- predict(rf2, data = test)
table(test$r2_AOO, pred$predictions)

rf2$variable.importance
```
