---
title: "SimulatedAtlasData"
author: "Friederike Wölke"
format: html
editor: visual
---

# Simulating Atlas data

The purpose of this script is to simulate simple Atlas data to test some of the hypotheses in a known environment and determine which techniques best recover the parameters that were used to simulate the data.

Steps:

1.  Read in Libraries

2.  Get the standard grid of Czechia & its' borders

3.  Simulate presence data for Czechia

### 1. Libraries

```{r}
rm(list=ls())
pacman::p_load(sparta, rnaturalearth, sf, raster, tidyverse, fractaldim)
```

### 2. Variables

```{r}
#| label: Variables 1 dataset
#| warning: false
#| echo: false
#| message: false

# folder path to Atlas data
source_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Datasets/Processed/Atlases/Replicated/Birds_Atlas_Czechia/"

# folder path to output folder
out_path <- "c:/Users/wolke/OneDrive - CZU v Praze/Dokumenty/GitHub/BEAST_General_Procedures/Scripts_Simulations/"

# create path to read in data and grids from variables
grid_path <-  paste0(source_path,"Birds_Atlas_Czechia_grid.gpkg") 

# save names of layers from file (needed to read them in):
layers <- st_layers(grid_path)$name

```

### 2. Create a grid layer of the Czech Republic (`POLYGONS`)

\-\-- Use data prepared by Gabriel

```{r}

#cells_corr <- readRDS(paste0(source_path, "Birds_Atlas_Czechia_cells_corr.rds"))

grid_list <- sapply(layers, function(i) {
  st_read(grid_path, paste(i), quiet = TRUE)
}, simplify = FALSE)

grid_cells_list <- list()
for (i in 1:length(grid_list)){
  Grid_cells <- grid_list[[i]] %>% 
  select(cell_label, cell_area_km2, cell_long, cell_lat)
  grid_cells_list[[i]] <- Grid_cells
}

total_area <- grid_cells_list[[1]] %>% mutate(total_area = sum(cell_area_km2)) %>% pull(total_area) %>% st_drop_geometry() %>% unique()

library(tmap)
qtm(grid_list[[1]], fill = "cell_grouping")
str(grid_list[[1]])
```

\-\-\-- Use standard data (skip for now)

```{r}
# # Borders of Czech Republic
# CZ_borders <- st_read('../data/CZE_adm0.gpkg')
# CZ_borders <- CZ_borders %>% select(WBINCOME, WBDEBT, Shape_Leng, Shape_Area)
# 
# # 100km2 grids of Czech Republic
# CZ_grids <- st_read('../data/KvadratyCR_JTSK.gpkg')
# 
# ## Do they have the same CRS? 
# st_crs(CZ_borders) == st_crs(CZ_grids) # no
# 
# # Transform the borders to have the same CRS so we can cut by the borders:
# CZ_borders <- st_transform(CZ_borders, crs = st_crs(CZ_grids)) # the same CRS as the CZ_grids layer
# 
# # Plot transformed:
# ggplot() + 
#   geom_sf(data=CZ_borders, fill='white') + 
#   geom_sf(data=CZ_grids, fill=NA) 
# 
# # Crop cells to borders:
# CZ_cropped <- st_intersection(CZ_grids, CZ_borders)
# 
# # Plot cropped map:
# ggplot() + 
#   geom_sf(data=CZ_cropped, fill='white')
# 
# # extract dataframe from grid to extract the details of the grid cells:
# ## We need this to simulate presence data based on the cells (OBJECTID)
# grid_cells_list[[1]] <- CZ_grids %>% select(OBJECTID, AREA, X, Y)
# 
# 
# # Remove columns that produce clutter:
# CZ_sf <- CZ_cropped %>%
#   select(OBJECTID, ENTITY, AREA, PERIMETE, X, Y, WBINCOME, WBDEBT)
```

### 3. Simulating presence data (`POINTS`)

```{r}
# extract parameters from Czechia map:

CZ_nSites <- length(unique(grid_list[[1]]$cell_label))
CZ_cell_label <- unique(grid_list[[1]]$cell_label)



# Create data 

## (this part was copied & modified from: https://biologicalrecordscentre.github.io/sparta/articles/sparta_vignette.html#modelling-methods)


n <- 300 # size of dataset
nyr <- 20 # number of years in data
nSamples <- 1000 # set number of dates
nSites <- CZ_nSites # set number of sites
set.seed(125) # set a random seed

# Create somes dates
first <- as.Date(strptime("1990/01/01", "%Y/%m/%d")) 
last <- as.Date(strptime(paste(1990+(nyr-1),"/12/31", sep=''), "%Y/%m/%d")) 
dt <- last-first 
rDates <- first + (runif(nSamples)*dt)

# taxa are set semi-randomly
#taxa_probabilities <- seq(from = 0.1, to = 0.7, length.out = 3) 
taxa_probabilities <- c(0.1, 0.5, 0.7)
verbatim_name <-  sample(letters[1:3], size = n, TRUE, prob = taxa_probabilities) # 3 taxa
table(verbatim_name)

# sites are visited semi-randomly
site_probabilities <- seq(from = 0.1, to = 0.7, length.out = nSites)
cell_label <- sample(CZ_cell_label, size = n, TRUE, prob = site_probabilities)

# the date of visit is selected semi-randomly from those created earlier
time_probabilities <- seq(from = 0.1, to = 0.7, length.out = nSamples)
time_period <- sample(rDates, size = n, TRUE, prob = time_probabilities)

myData <- data.frame(verbatim_name, cell_label, time_period)

# Let's have a look at the my example data
head(myData)

## Create a new column for the time period
# First define my time periods
time_periods <- data.frame(start = c(1990, 2000),
                           end = c(1999, 2009))

# Now use these to assign my dates to time periods
myData$tp <- sparta::date2timeperiod(myData$time_period, time_periods)
myData$tp <- as.factor(myData$tp)
```

#### Merge Presence data with CZ grid

```{r}
my_Data <- merge(myData,grid_cells_list[[1]])
myData_sf <- my_Data %>% 
  filter(!is.na(cell_long) & !is.na(cell_lat)) %>% # filter records without coordinates
  st_as_sf(coords=c('cell_long', 'cell_lat'),
           crs=st_crs(grid_cells_list[[1]]))

myData_sf_sub <- myData_sf %>% select(verbatim_name, tp)
```

#### Aggregate fine-grain data to bigger grains

```{r}
pres_dat_scaled <- list()

for (i in 1:length(grid_cells_list)){
  temp_group <- st_join(grid_cells_list[[i]], myData_sf_sub)
  pres_dat_scaled[[i]] <- temp_group
  }

```

#### Use SPARTA package to assess the quality of the data

```{r}
# head(pres_dat_scaled[[1]])
# table(pres_dat_scaled[[1]]$verbatim_name)
# length(is.na(pres_dat_scaled[[1]]$verbatim_name))
# 
# ## Now let's use the SPARTA package to assess the quality of our data:
# results <- dataDiagnostics(taxa = myData_sf_sub$verbatim_name,
#                            site = rownames(myData_sf_sub),
#                            time_period = myData_sf_sub$tp,
#                            progress_bar = FALSE)
# 
# 
# # Calculate relative change with Telfer index:
# telfer_results <- telfer(taxa = myData_sf_sub$verbatim_name,
#                          site = rownames(myData_sf_sub),
#                          time_period = myData_sf_sub$tp,
#                          minSite = 1)
# 
# head(telfer_results)
```

#### Exploring the simulated data

```{r}

# 2. Plot occurrences per species:

p_list <- list()
occ_list <- list()
occ_list_scaled <- list()

for (j in 1:length(pres_dat_scaled)){
  curr_grain <- pres_dat_scaled[[j]] %>% arrange(verbatim_name) %>% unique()
  rownames(curr_grain) <- NULL
  curr_grain$total_area <- total_area
  
  
  ## Calculate Species Richness:
  Presence_data_SR1 <- curr_grain %>% filter(tp == 1) %>%
  group_by(cell_label) %>% unique() %>%
  summarise(N=sum(!is.na(verbatim_name)), 
            SR=n_distinct(verbatim_name, na.rm = TRUE))
  
  
  Presence_data_SR2 <- curr_grain %>% filter(tp == 2) %>%
  group_by(cell_label)  %>% unique() %>% 
  summarise(N=sum(!is.na(verbatim_name)), 
            SR=n_distinct(verbatim_name, na.rm = TRUE))

  # 1. Plot Species Richness:
  # p <- ggplot() + 
  #   geom_sf(data=Presence_data_SR, aes(fill=SR)) +
  #   scale_fill_fermenter(palette ='YlOrBr', n.breaks=3, direction = 1) 
  # p_list[[j]] <- p
  
  qtm(Presence_data_SR1, fill = "SR")
  qtm(Presence_data_SR2, fill = "SR")
  
  
  for (i in 1:length(unique(curr_grain$verbatim_name))){
  
    taxon <- unique(curr_grain$verbatim_name)[i]
    temp <- curr_grain %>% filter(verbatim_name == taxon)
    
    # Now calculate fine-grain occupancy:
    occ_data <- temp %>% st_drop_geometry() %>%
      group_by(tp) %>% unique() %>%
      mutate(occupancy_area = sum(cell_area_km2)) %>%  
      mutate(relative_occupancy = occupancy_area/total_area) %>%
      mutate(relative_occupancy = round(relative_occupancy,2)) %>% 
      filter(!is.na(verbatim_name)) %>%  
      ungroup() %>%
      select(verbatim_name, tp, occupancy_area, relative_occupancy, cell_area_km2, total_area) %>%
      unique() 
    
  
  
  occ_list[[i]] <- occ_data
  
  }
  occ_df <- do.call(rbind, occ_list)
  occ_df$cell_grouping <- j
  occ_list_scaled[[j]] <- occ_df 
  
}

# get an overview by plotting the first 10 results (includes NA)
p_list

occ_df_scaled <- do.call(rbind,occ_list_scaled)
```

```{r}
temp_df <- occ_df_scaled %>% filter(tp == 1, verbatim_name == "a") %>% unique()

D_a1 <- summary(lm(log10(relative_occupancy)~log10(cell_area_km2 ), data=temp_df))$coef[2,1]


temp_df <- occ_df_scaled %>% filter(tp == 1, verbatim_name == "b") %>% unique()

D_b1 <- summary(lm(log10(relative_occupancy)~log10(cell_area_km2 ), data=temp_df))$coef[2,1]


temp_df <- occ_df_scaled %>% filter(tp == 1, verbatim_name == "c") %>% unique()

D_c1 <- summary(lm(log10(relative_occupancy)~log10(cell_area_km2 ), data=temp_df))$coef[2,1]



temp_df <- occ_df_scaled %>% filter(tp == 2, verbatim_name == "a") %>% unique()

D_a2 <- summary(lm(log10(relative_occupancy)~log10(cell_area_km2 ), data=temp_df))$coef[2,1]


temp_df <- occ_df_scaled %>% filter(tp == 2, verbatim_name == "b") %>% unique()

D_b2 <- summary(lm(log10(relative_occupancy)~log10(cell_area_km2 ), data=temp_df))$coef[2,1]


temp_df <- occ_df_scaled %>% filter(tp == 2, verbatim_name == "c") %>% unique()

D_c2 <- summary(lm(log10(relative_occupancy)~log10(cell_area_km2 ), data=temp_df))$coef[2,1]

D1 <- c(D_a1, D_b1, D_c1)
D2 <- c(D_a2, D_b2, D_c2)
D1D2 <- cbind(D1,D2)
rownames(D1D2) <- c("a", "b", "c")
D1D2
```
